Does an earthquake “know” how big it will be? A neural-net aided study

### 2024-03-15 article.tex
Modify into, update figures slightly

\title[Running Title]{Does an earthquake “know” how big it’s going to be? A neural-net aided study} % 2nd option: Using past seismicity to predict the magnitude of future earthquakes

@@

Earthquake forecasting is a long-standing scientific and technological challenge, often deemed unrealistic due to the inherent complexity of earthquake processes \cite{bernard_earthquake_1999, geller_earthquakes_1997}. Research since the late 19th century has provided much phenomenological insight about the spatiotemporal statistics of earthquakes. This includes various characteristics of spatial and temporal clustering \cite{omori_after-shocks_1894, kagan_short-term_2004, ben-zion_localization_2020}, which can be incorporated into stochastic prediction schemes \cite{ogata_statistical_1988, hardebeck_aftershock_2024, devries_deep_2018, king_static_1994}. These forecasting methods focus on the rate, and sometimes the location, of earthquake occurrence. Furthermore, these spatiotemporal properties are almost always treated independently from the magnitude, an assumption known as separability \cite{schoenberg_testing_2004}. That is, while seismic history is used to inform the location and rate of future earthquakes, i.e. by assuming some phenomenological description of spatiotemporal clustering, the magnitude of future events prediction often relies on drawing values from the Gutenberg-Richter (GR) distribution or its variants \cite{gutenberg_frequency_1944, kagan_seismic_2002, ogata_exploring_2018}. Importantly, this distribution does not depend on the seismic history, implying that, conditioned on the fact that an earthquake occurs, its magnitude is taken to be independent of the rate and location of past events.

@@

While assuming that an earthquake's magnitude is independent of the seismic history is popular sue to its simplicity and its ease of use in earthquake modeling, it is not necessarily true. Indeed, the magnitude of an earthquake is a property of the rupture process, which is governed by the frictional properties of the fault and the stress field. These in turn depend on slip history in a deterministic manner, albeit intangibly complicated. Therefore, the assumption of magnitude separability, i.e. its independent from seismic history, is not a priory justified, and its validity should be tested empirically.
Though information about an earthquake magnitude may exists before its initiation, implying dependence on seismic history and non-separability, it has yet to be shown accessible using our current experimental or analytic tools. Current research on magnitude prediction has failed to show a universal and replicable advantage over the common benchmarks \cite{shcherbakov_forecasting_2019, ogata_exploring_2018, stockman_forecasting_2023,  panakkat_neural_2007}, by that leaving the question of accessible information prior to an earthquake's nucleation withstanding. 
\newline
In this paper we explore if information about an earthquake's magnitude can be obtained from regional seismic history. This is done by constructing a neural-based model that predicts the magnitude of a given earthquake, given the short and long term seismic history prior to its occurrence. Importantly, our model is not tasked with predicting the timing and location of the event, as they are explicitly provided to the model. Thus, we separate out the task of predicting the nucleation of events and isolate the question of magnitude predictability. If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of an earthquake is extractable from cataloged properties alone.

This contribution has two important consequences. From a fundamental point of view, it challenges the self-organized-criticality viewpoint of earthquake magnitude emergence, which posits that magnitudes are inherently unpredictable \cite{olami_self-organized_1992, sornette_self-organized_1989, bak_earthquakes_1989, de_geus_scaling_2022}. Second, it suggests that the separability assumption, which is widely applied in earthquake forecasting, may be replaced by a more nuanced model that incorporates the seismic history into the magnitude prediction. This may lead to improved forecasting models, and potentially also be used to identify precursory signals in the seismic history of an earthquake.

@@

We construct a neural network (NN) model that encodes the seismic history of time series from a hypocentral catalog and produces an estimate of the possible magnitudes at a queried time and place. The constructed NN uses a long-short term memory (LSTM) architecture to model the seismic history up to a given time. This latent representation is then passed together with  a space-time coordinate query of a future earthquake into a fully connected neural network (FCNN) that produces a parametrized probability density function (PDF) of the possible magnitudes at the quarried space- time coordinates. This procedure is depicted in Fig. \ref{fig:intro_fig}a depicts the task as described here. A detailed description of the model's architecture is given in the methods section. We optimize the model parameters to maximize the log likelihood (LL) of the true magnitude label, by using the loss function $\mathcal{L}$:
\begin{equation}
    \mathcal{L} = -\langle \log{ \left( p_{\textbf{x}_i, t_i} \left( m_i \right) \right) } \rangle\ ,
    \label{eq:loss_function}
\end{equation}
where $p_{\textbf{x}_i, t_i}$ is the probability function returned by the model for the space-time coordinates $\textbf{x}_i, t_i$ (i.e. the ``query coordinates''), $m_i$ the corresponding to the actual magnitude of the $i^{th}$ earthquake and $\langle \cdot\rangle$ stands for the empirical average over the data set. We use the a mixture of two Kumaraswamy distributions \cite{kumaraswamy_generalized_1980} parameterized family of probability distributions stretched to a relevant domain. The final distribution used is defined by 5 parameters. This family smoothly can interpolate between a decaying distribution (resembling the GR distribution), and localized distributions whose mass is concentrated around a specific value. This allows the model to output both an ``ignorant'' prediction, essentially resembling the GR distribution, and more confident predictions localized around a given magnitude.

We train the model on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: Hauksson et al. (cite) for Southern California, GeoNet (cite) for New Zealand, and the JMA catalog (cite) for [Region 1, Region 2...]. While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies (e.g.? cite?) and exhibit varying data quality. A separate model is trained for each region, with identical loss function and parameterization of the PDF.

Model performance is evaluated over a time span that was not included in training. The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating an information gain in forecasting earthquake magnitudes prior to their occurrence. \neri{I think this is a good place to add that we conclude that there is some preparatory phase.}
%This finding directly suggests that the seismic system undergoes preparatory processes tailored to the impending event's magnitude before its initiation. Examining these results across multiple catalogs and regions highlights the robustness of our methodology and the generalizability of our conclusion, paving the way for further exploration of earthquake predictability.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/raw_results_hauksson.pdf}
    \caption{
        Caption.
    }
    \label{fig:architecture}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \fbox{
        \rule{1\textwidth}{0pt}\rule{0pt}{8ex}
        FIGURE OF SPATIAL DISTRIBUTION OF GAIN.
    }
    \caption{
        Caption.
    }
    \label{fig:architecture}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.4\textwidth]{figures/ROC_PR_hauksson.pdf}
    \caption{
        Caption.
    }
    \label{fig:architecture}
\end{figure}

>>>>

\title[Running Title]{Does an earthquake “know” how big it will be? A neural-net aided study} % 2nd option: Using past seismicity to predict the magnitude of future earthquakes

@@

Earthquake forecasting is a long-standing scientific and technological challenge, often deemed unrealistic due to the inherent complexity of earthquake processes \cite{bernard_earthquake_1999, geller_earthquakes_1997}. Research since the late 19th century has provided much phenomenological insight about the spatiotemporal statistics of earthquakes. This includes various characteristics of spatial and temporal clustering \cite{omori_after-shocks_1894, kagan_short-term_2004, ben-zion_localization_2020}, which can be incorporated into stochastic prediction schemes \cite{ogata_statistical_1988, hardebeck_aftershock_2024, devries_deep_2018, king_static_1994}. These forecasting methods focus on the rate, and sometimes the location, of earthquake occurrence [cite]. Furthermore, these spatiotemporal properties are almost always treated independently from the magnitude, an assumption known as separability \cite{schoenberg_testing_2004}. That is, while seismic history is used to inform the location and rate of future earthquakes, i.e. by assuming some phenomenological description of spatiotemporal clustering, the magnitude of future events prediction often relies on drawing values from the Gutenberg-Richter (GR) distribution or its variants \cite{gutenberg_frequency_1944, kagan_seismic_2002, ogata_exploring_2018}. Importantly, this distribution either depends weakly or does not depend on the seismic history, implying that, conditioned on the fact that an earthquake occurs, its magnitude is taken to be independent of the rate and location of past events.

@@

While assuming that an earthquake's magnitude is independent of the seismic history is popular due to its simplicity and its ease of use in earthquake modeling, it is not necessarily true. Indeed, the magnitude of an earthquake is a property of the rupture process, which is governed by the frictional properties of the fault and the stress field. These in turn depend on slip history in a deterministic manner, albeit intangibly complicated. Therefore, the assumption of magnitude separability, i.e. its independent from seismic history, is not a priory justified, and its validity should be tested empirically.
Though information about an earthquake magnitude may exists before its initiation, implying dependence on seismic history and non-separability, it has yet to be shown accessible using our current experimental or analytic tools. Current research on magnitude prediction has failed to show a universal and replicable advantage over the common benchmarks \cite{shcherbakov_forecasting_2019, ogata_exploring_2018, stockman_forecasting_2023,  panakkat_neural_2007}, by that leaving the question of accessible information prior to an earthquake's nucleation withstanding. 
\newline
% perhaps a good place to mention Giuseppe-Zhuang work, Sam's  and Kelian's
In this paper we explore if information about an earthquake's magnitude can be obtained from regional seismic history. This is done by constructing a neural-based model that predicts the magnitude of a given earthquake, given the short and long term seismic history prior to its occurrence. Importantly, our model is not tasked with predicting the timing and location of the event, as they are explicitly provided to the model. Thus, we separate out the task of predicting the nucleation of events and isolate the question of magnitude predictability. If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of an earthquake is extractable from cataloged properties alone.

This contribution has two important consequences. From a fundamental point of view, it challenges the self-organized-criticality viewpoint of earthquake magnitude emergence, which posits that magnitudes are inherently unpredictable and do not depend on seismic history \cite{olami_self-organized_1992, sornette_self-organized_1989, bak_earthquakes_1989, de_geus_scaling_2022}. Second, it suggests that the separability assumption, which is widely applied in earthquake forecasting, may be replaced by a more nuanced model that incorporates the seismic history into the magnitude prediction. This may lead to improved forecasting models, and potentially also be used to identify precursory signals in the seismic history of an earthquake.

@@

In order to tackle the question of magnitude predictability, we construct the MAGnitude prediction Neural NETwork model, MAGNNET, a neural network (NN) for producing a probability density function of the possible magnitudes at a queried space time coordinates, conditioned on the seismic history.

MAGNNET encodes the seismic history of time series from a hypocentral catalog and produces an estimate of the possible magnitudes at a queried time and place.

MAGNNET's result will be quantitatively compared to other known benchmarks to estimate the amount of information gained by encoding the seismic history.
MAGNNET is composed of a long-short term memory (LSTM) architecture to model the seismic history up to a given time. This latent representation is then passed together with  a space-time coordinate query of a future earthquake into a fully connected neural network (FCNN) that produces a parametrized probability density function (PDF) of the possible magnitudes at the quarried space- time coordinates. This procedure is depicted in Fig. \ref{fig:intro_fig}a. A detailed description of the model's architecture is given in the methods section \neri{TODO}. We optimize MAGNNET's parameters to maximize the log likelihood (LL) of the true magnitude label, by using the loss function $\mathcal{L}$:
\begin{equation}
    \mathcal{L} = -\langle \log{ \left( p_{\textbf{x}_i, t_i} \left( m_i \right) \right) } \rangle\ ,
    \label{eq:loss_function}
\end{equation}
where $p_{\textbf{x}_i, t_i}$ is the probability function returned by the model for the space-time coordinates $\textbf{x}_i, t_i$ (i.e. the ``query coordinates''), $m_i$ the corresponding to the actual magnitude of the $i^{th}$ earthquake and $\langle \cdot\rangle$ stands for the empirical average over the data set. We use the a mixture of two Kumaraswamy distributions \cite{kumaraswamy_generalized_1980} parameterized family of probability distributions stretched to a relevant domain. The final distribution used is defined by 5 parameters. This family smoothly can interpolate between a decaying distribution (resembling the GR distribution), and localized distributions whose mass is concentrated around a specific value. This allows the model to output both an ``ignorant'' prediction, essentially resembling the GR distribution, and more confident predictions localized around a given magnitude.

We train MAGNNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog [cite] for Southern California, GeoNet [cite] for New Zealand, and the JMA catalog [cite] for Japan. While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies (e.g.? cite?) and exhibit varying data quality. A separate model is trained for each region, with identical loss function and parameterization of the PDF.

The model's, MAGNNET's, performance is evaluated over a time span that was not included in training. The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating an information gain in forecasting earthquake magnitudes prior to their occurrence. \neri{I think this is a good place to add that we conclude that there is some preparatory phase.}
%This finding directly suggests that the seismic system undergoes preparatory processes tailored to the impending event's magnitude before its initiation. Examining these results across multiple catalogs and regions highlights the robustness of our methodology and the generalizability of our conclusion, paving the way for further exploration of earthquake predictability.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/raw_results_hauksson.pdf}
    \caption{
        \textbf{outputs of MAGnitude prediction Neural NETwork, MAGNNET, for the Hauksson catalog of Southern California. a}, Likelihoods of MAGNNET and some common benchmark magnitude predictors. Solid lines represent the median value along the magnitude axis, shading shows the 10th and 90th percentiles of the distribution along the vertical axis. \textbf{b}, Train set's Gutenberg-Richter dostribution (dashen black line) with PDFs produced by MAGNNET for 100 randomly sampled events from the test set. Each PDF is color coded by the label of the event, its true magnitude. Colorbar to the right indicates the color code. \textbf{c}, Total magnitude frequency distribution produced by MAGNNET and common benchmarks for the entire test set. Histograms of the train (and test) set are presented in orange (blue).
    }
    \label{fig:architecture}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/combined_batplots.pdf}
    \caption{
        \textbf{Metric scores for out model MAGNNET, and common benchmark models. a}, \textbf{b}, \textbf{c}, Minus mean information content of our MAGNNET model (red) and other common benchmark magnitude predictors (see labels on bars in the figure). \textbf{d, e, f}, Reciever Operating Characteristic (ROC) curve for a binary classifier determinng the existence of a large ($m>=4$) event. Area under the curve (AUC) for each curve is written in each frame, colorcoded identically to the bar plots. \textbf{g, h, i}, Precision at recall curves for the binary classifier described in d, e, f. 
    }
    \label{fig:architecture}
\end{figure}

@@
